Thermal reactionomes reveal adaptive responses to thermal extremes in warm

### 2014-10-24 doc/ApTranscriptome_manuscript.Rmd

---
title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: pnas1.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhancing the cellular stress response (*enhanced defense*), consititutively elevated expression of protective genes (*genetic assimilation*) or shifting from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperature intervals that spanned their entire thermal breadth. We found that at least 8% of all genes were thermally responsive. The majority of upregulation was specific to low temperatures, and consistent with the *enhanced defense hypothesis* the cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures. We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression. In support of the *tolerance hypothesis* but not the *enhanced defense hypothesis*, transcripts turned on at high temperatures were induced on average at `r round(T_on.high.ttest$estimate[1], 1)`°C in *A. carolinensis* compared to `r round(T_on.high.ttest$estimate[2], 1)`°C in *A. picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For low temperatures, there was no difference in the mean temperature of transcript upregulation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

Similarly, for the **Intermediate** transcripts we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures. Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately -`r round(mean(T_on.high.ttest$estimate) - mean(T_off.int.high.ttest$estimate), 1)`°C at the higher temperature extremes and a shift of `r round(mean(T_on.low.ttest$estimate) - mean(T_off.int.low.ttest$estimate), 1)`°C at low temperature extremes. There was greater inertia of gene expression in response to high and low temperature extremes. With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), whereas with decreasing temperatures **Intermediate** transcripts from *A. picea* were downregulated at higher temperatures (`r round(T_off.int.low.ttest$estimate[2], 1)`) than those from *A. carolinensis* (`r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).

@@

To test for a tradeoff between induciblity and constitutive baseline expression between species, we fit ordinary least squares regressions with the log ratio of relative constitutive expression as the response and the log ratio of relative inducibility as the predictor for **High** transcripts in *A. picea* and for **Low** transcripts in *A. carolinensis*. Constitutive expression was defined as predicted expression at 19.5°C, while inducibility of each transcript was defined as *(maximum TPM - minimum TPM) / minimum TPM * 100*.

@@

Figure legends go here.

>>>>

---
title: Thermal reactionomes reveal adaptive responses to thermal extremes in warm
  and cool-climate ant species
author: ' '
csl: pnas1.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: no
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

The distributions of species and their responses to climate change are in part determined by their thermal tolerances. However, little is known about how thermal tolerance evolves. To test whether evolutionary extension of thermal limits is accomplished through enhanced cellular stress response (*enhanced defense*), constitutively elevated expression of protective genes (*genetic assimilation*) or a shift from damage resistance to passive mechanisms of thermal stability (*tolerance*), we conducted an analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized thermal reactionomes of two common ant species in the eastern U.S, the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*, across 12 temperatures that spanned their entire thermal breadth. We found that at least 8% of all genes were thermally responsive. The majority of upregulation was specific to low temperatures, and consistent with *enhanced defense* the cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins. In contrast, under high temperatures the warm-adapted *A. carolinensis* downregulated many of the genes upregulated in *A. picea* and required more extreme temperatures to induce changes in gene expression, consistent with the *tolerance* hypothesis. We found no evidence for a trade-off between constitutive and inducible gene expression as predicted by the *genetic assimilation hypothesis*. These results are consistent with expectations that adaptive increases in upper thermal limits may require an evolutionary shift in response mechanism away from damage repair toward tolerance and prevention.

@@

As an additional test of the *tolerance hypothesis*, we examined the critical temperature of gene induction in response to increasing and decreasing temperatures. We compared between species the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression. In support of the *tolerance hypothesis* but not the *enhanced defense hypothesis*, transcripts turned on at high temperatures were induced on average at `r round(T_on.high.ttest$estimate[1], 1)`°C in *A. carolinensis* compared to `r round(T_on.high.ttest$estimate[2], 1)`°C in *A. picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2A). For low temperatures, there was no difference in the mean temperature of transcript upregulation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2A).

Similarly, for the **Intermediate** transcripts we compared the mean temperatures of downregulation of transcript expression between species at both high (> 20°C) and low (< 20°C) temperatures. Thresholds for downregulation were shifted toward cooler temperatures than were the corresponding thresholds for gene activation, with a shift of approximately -`r round(mean(T_on.high.ttest$estimate) - mean(T_off.int.high.ttest$estimate), 1)`°C at the higher temperature extremes and a shift of `r round(mean(T_on.low.ttest$estimate) - mean(T_off.int.low.ttest$estimate), 1)`°C at low temperature extremes (Fig. 2). There was greater inertia of gene expression in response to high and low temperature extremes. With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), whereas with decreasing temperatures **Intermediate** transcripts from *A. picea* were downregulated at higher temperatures (`r round(T_off.int.low.ttest$estimate[2], 1)`) than those from *A. carolinensis* (`r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* = `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`, Fig. 2B).

@@

To test for a tradeoff between induciblity and constitutive baseline expression between species, we fit ordinary least squares regressions with the log ratio of relative constitutive expression as the response and the log ratio of relative inducibility as the predictor for **High** transcripts in *A. picea* and for **Low** transcripts in *A. carolinensis*. Constitutive expression was defined as predicted expression at 19.5°C, while inducibility of each transcript was defined as *(maximum TPM - minimum TPM) / minimum TPM * 100*. In addition, we used a Mann-Whitney test to compare the baseline constitutive expression between species for all responsive transcripts.

@@

**Figure 1**. Deviations from expected numbers of transcripts in matched observations of transcript expression type. The color of each cell represents the deviation of the observed from the expected number of transcripts based on hypothetical equivalence of the marginal frequencies (Blue = excess, Red = deficit). The dashed boxes indicate the diagonals (e.g. **High** - **High**). (A) **Low** and **Bimodal** expression types for *A. carolinensis* showing bias away from all other categories in *A. picea*. (B) **High** and **Bimodal** expression types for *A. picea* showing excess of transcripts that are not upregulated (**Intermediate** and **Not Responsive**). The complete set of matched observations is shown in Fig. S1.

**Figure 2**. Histogram with smooth density estimate of temperature of maximum change in expression for transcripts that have increased expression at **Low** and **High** temperatures. Red bars and lines are for *A. carolinensis* while blue bars and lines are for *A. picea*.

**Figure 3**. Scatterplots of log ratios of relative inducibility to relative constitutive expression for (A) **High** transcripts in *A. picea* (*P* < 0.001, r^2^ = 0.07) and (B) **Low** transcripts in *A. carolinensis* (*P* < 0.001, r^2^ = 0.1). Blue lines and confidence intervals are from ordinary least squares regressions.

# Tables

# Figures
